Short-time simulation of open collective quantum spin systems

### 2018-11-25 soc_squeezing/collective_spin.tex
made intro text less specific to squeezing/OAT slightly simplified expectation value expressions

\documentclass[aps,11pt,notitlepage,nofootinbib]{revtex4-1}

@@

Collective spin systems attract considerable attention for providing a
window into many-body quantum phenomena that are both experimentally
accessible\cite{fernholz2008spin, takano2009spin,
  riedel2010atomchipbased, gross2010nonlinear, martin2013quantum,
  bohnet2016quantum, hosten2016quantum, norcia2018cavitymediated} and
metrologically useful\cite{wineland1992spin, kitagawa1993squeezed,
  ma2011quantum}.  Theoretical interest in such systems dates back to
the mid-twentieth century with the introduction of the
Lipkin-Meshkov-Glick (LMG) model as a toy model for testing many-body
approximation methods in contemporary nuclear
physics\cite{lipkin1965validity, meshkov1965validity,
  glick1965validity}.  Since then, collective system systems have
additionally featured in theoretical studies of quantum
criticality\cite{latorre2005entanglement, alcalde2007functional,
  morrison2008collective, sarandy2009classical, wang2012quantum,
  majd2014lmg}, non-equilibrium
phenomena\cite{morrison2008dissipationdriven,
  bhattacherjee2014nonequilibrium, klinder2015dynamical,
  zhiqiang2017nonequilibrium, smale2018observation}, and most notably
spin squeezing\cite{wineland1992spin, kitagawa1993squeezed,
  ma2011quantum}.  With the development of advanced atomic trapping,
cooling, and control techniques, collective spin systems have been
experimentally realized in atomic ensembles\cite{fernholz2008spin,
  takano2009spin}, Bose-Einstein
condensates\cite{riedel2010atomchipbased, gross2010nonlinear},
ultracold Fermi gasses\cite{martin2013quantum}, trapped
ions\cite{bohnet2016quantum}, and optical
cavities\cite{hosten2016quantum, norcia2018cavitymediated}.

One of the primary current motivations for studying collective spin
systems is their capability to dynamically generate spin-squeezed
states, which enable measurement of phase angles $\phi$ with an error
$\Delta\phi$ below the standard quantum limit
$\Delta\phi\sim1/\sqrt{N}$ set by quantum projection noise from the
measurement of $N$ independent particles\cite{ma2011quantum}.  The
nonlinear dynamics necessary to generate spin-squeezed states can be
implemented with long-range collisional, photon-, or phonon-mediated
interactions, typically resulting in a one-axis twising (OAT) model
described by the Hamiltonian\cite{kitagawa1993squeezed}
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
  \label{eq:OAT}
\end{align}
where $S_\z\equiv\sum_{j=1}^N\sigma_\z^{(j)}/2$ is a collective
spin-$z$ operator defined in terms of Pauli-$z$ operators
$\sigma_\z^{(j)}$ addressing spin $j$.  The OAT model prepares a
spin-squeezed state that allows measurement of a phase angle $\phi$
with error $\Delta\phi\sim N^{-2/3}$ at times
$t\sim\chi^{-1}N^{-2/3}$\cite{ma2011quantum}.  Although a truly
collective spin model requires uniform, all-to-all interactions, as
long as measurements do not distinguish between constituent particles
even non-uniform systems can be effectively described by a uniform
model with renormalized parameters\cite{hu2015entangled}.  Various
works have proposed improvements over OAT through the addition of
auxiliary control fields\cite{law2001coherent, liu2011spin,
  huang2015twoaxis, muessel2015twistandturn} or a modified
vacuum\cite{hu2017vacuum}, thereby realizing different collective spin
models.

While the OAT model admits exact
solutions\cite{foss-feig2013nonequilibrium}, the same is not true of
more general, non-Ising Hamiltonians.  Nonetheless, in the absence of
decoherence, permutation symmetry and total spin conservation divide
the total Hilbert space of a collective spin system into superseletion
sectors that grow only linearly with system size, thereby admitting
efficient classical simulation of its dynamics.  If decoherence is
present, but sufficiently weak, dynamics of collective spin systems
are still numerically solvable via ``quantum jump'' Monte Carlo
methods\cite{plenio1998quantumjump, zhang2018montecarlo} (also known
as ``quantum trajectory'' or ``Monte Carlo wavefunction'' methods)
that can reproduce all expectation values of interest.  When
decoherence is strong, however, or when the number of jump operators
grows extensively with system size (e.g.~for single-spin decay), these
Monte Carlo methods can take a prohibitively long time to converge, as
simulations become dominated by incoherent jumps that generate large
numbers of distinct trajectories over which to average.

In this work, we develop a new method to simulate short-time dynamics
of collective spin systems with permutationally-symmetric decoherence.
This method is based on a short-time expansion of exact solutions to
the equations of motion for Heisenberg operators.  Evaluating this
expansion requires use of the structure factors of a collective spin
operator algebra; the calculation of these structure factors is one of
the main technical results of this work.

@@

The time derivative operator $d/dt$ will generally couple spin
operators $\S_{\v m}$ to spin operators $\S_{\v n}$ with higher
``weight'', i.e.~with $\abs{\v n}>\abs{\v m}$, where
$\abs{\v\ell}\equiv\sum_\lambda\ell_\lambda$.  The growth of operator
weight signifies the growth of many-body correlations.  In practice,
keeping track of this growth will eventually require more
computational resources than are available, meaning we must somehow
truncate our equations of motion.  The simplest truncation strategy
would be to take
\begin{align}
  \f{d}{dt} \S_{\v m}
  \to \sum_{w\p{\v n}<W} T_{\v m\v n} \S_{\v n}
  \label{eq:weight_truncation}
\end{align}
for some weight measure $w$, e.g.~$w\p{\v n}=\abs{\v n}$, and cutoff
$W$.  The truncation in Eq.~\eqref{eq:weight_truncation} will close
the system of differential equations defined by Eq.~\eqref{eq:EOM},
and allow us to solve it using standard numerical methods.  Initial
conditions for this system of differential equations, i.e.~expectation
values of collective spin operators with respect to states that are
simple to prepare experimentally, are provided in Appendix
\ref{sec:initial_conditions}.

The truncation strategy in Eq.~\eqref{eq:weight_truncation} has a few
limitations:
\begin{enumerate*}[label=(\roman*)]
\item simulating a system of differential equations for a large number
  of operators can be quite time-consuming,
\item the weight measure $w$ may need to be chosen carefully, as the
  optimal measure is generally system-dependent,
\item simulation results can only be trusted up to the time at which
  the initial value of operators $\S_{\v m}$ with weight
  $w\p{\v m}\ge W$ have a non-negligible contribution to operators of
  interest.\label{pt:limitation}
\end{enumerate*}
The last limitation in particular unavoidably applies in some form to
any method tracking only a subset of all relevant operators.  We
therefore devise an alternate truncation strategy built around
limitation \ref{pt:limitation}.

@@

\section{Benchmarking}

@@

Initial conditions for all collective spin operators with respect to
(spin-polarized) Gaussian states in the Dicke manifold spanned by
$\set{\ket{m}:S_\z\ket{m}=m\ket{m}}$ can be determined from the
expansions
\begin{align}
  S_\z = \sum_{k=-S}^S k \op{k},
  &&
  S_\mu = \sum_{k=-S+\delta_{\mu,-1}}^{S-\delta_{\mu,1}}
  g_\mu\p{S,k} \op{k+\mu}{k},
\end{align}
with $\mu\in\set{+1,-1}$ and
\begin{align}
  g_\mu\p{S,k} \equiv \sqrt{\p{S-\mu k}\p{S+\mu k+1}}.
\end{align}
Given the initial states
\begin{align}
  \ket{\pm\Z} \equiv \ket{\pm S},
  &&
  \ket{\X} \equiv 2^{-N/2} \sum_{k=-S}^S
  { N \choose S+k }^{1/2} \ket{k},
\end{align}
where $S_\x\ket{\X}=S\ket{\X}$, in terms of $\nu\in\set{+1,-1}$ and
$\bmu\equiv-\mu$ we then have
\begin{align}
  \bk{\nu\Z|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\nu\Z}
  = \delta_{\ell n} \times
  \begin{cases}
     \p{S-n}^m \f{N! n!}{\p{N-n}!} & \mu = \nu, \\
     \delta_{n,0} \p{-S}^m & \mu \ne \nu,
  \end{cases}
\end{align}
1and
\begin{align}
  \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}
  &=  \f{\mu^m}{2^N} \sum_{k=-S+\delta_{\mu,-1}\max\set{\ell,n}}
  ^{S-\delta_{\mu,1}\max\set{\ell,n}} k^m
  \sp{{ N \choose S+k+\mu\ell } { N \choose S+k+\mu n }}^{1/2}
  \notag \\
  &\qquad\qquad\qquad\qquad
  \times \prod_{p=0}^{\ell-1} g_\mu\p{S,k+\mu p}
  \prod_{q=0}^{n-1} g_\bmu\p{S,k+\mu\sp{n-q}} \\
  &= \mu^m \f{N!}{2^N} \sum_{k=-S+\delta_{\mu,-1}\max\set{\ell,n}}
  ^{S-\delta_{\mu,1}\max\set{\ell,n}}
  \f{k^m \p{S-\mu k}!}
  {\p{S+\mu k}!\p{S-\ell-\mu k}!\p{S-n-\mu k}!} \\
  &= \f{N!}{2^N} \sum_{k=-S}^{S-\max\set{\ell,n}}
  \f{k^m \p{S-k}!}{\p{S+k}!\p{S-\ell-k}!\p{S-n-k}!} \\
  &= \p{-1}^m\f{N!}{2^N} \sum_{k=0}^{N-\max\set{\ell,n}}
  \f{\p{S-k}^m \p{N-k}!}{k!\p{N-\ell-k}!\p{N-n-k}!}.
\end{align}
Furthermore, using the fact that
\begin{align}
  e^{i\phi S_\z} S_\mu e^{-i\phi S_\z} = e^{i\mu\phi} S_\mu,
\end{align}
we can define the state
\begin{align}
  \ket{\tilde\phi} \equiv e^{-i\phi S_\z}\ket{\X},
  &&
  \p{\cos\phi S_\x + \sin\phi S_\y} \ket{\tilde\phi}
  = S \ket{\tilde\phi},
\end{align}
and say that
\begin{align}
  \bk{\tilde\phi|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\tilde\phi}
  = e^{i\mu\p{\ell-n}\phi} \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}.
\end{align}

@@

$P_{\v j}$ in \eqref{eq:SS_P} we define

>>>>

\documentclass[aps,11pt,notitlepage,nofootinbib,longbibliography]{revtex4-1}

@@

Collective spin systems attract considerable attention for providing a
window into many-body quantum phenomena that are both experimentally
accessible\cite{takano2009spin, chen2010heralded,
  riedel2010atomchipbased, gross2010nonlinear, martin2013quantum,
  bohnet2016quantum, hosten2016quantum, norcia2018cavitymediated,
  jones2009magnetic} and metrologically
useful\cite{riedel2010atomchipbased, gross2010nonlinear,
  hosten2016quantum, jones2009magnetic, zhong2010simplified,
  huang2015quantum, agarwal1997atomic, lau2014proposal,
  wineland1992spin, kitagawa1993squeezed, ma2011quantum}.  Theoretical
interest in such systems dates back to the mid-twentieth century with
the introduction of the Lipkin-Meshkov-Glick (LMG) model as a toy
model for testing many-body approximation methods in contemporary
nuclear physics\cite{lipkin1965validity, meshkov1965validity,
  glick1965validity}.  Since then, collective system systems have
additionally featured in theoretical studies of quantum
criticality\cite{latorre2005entanglement, alcalde2007functional,
  morrison2008collective, sarandy2009classical, wang2012quantum,
  majd2014lmg}, non-equilibrium
phenomena\cite{morrison2008dissipationdriven,
  bhattacherjee2014nonequilibrium, klinder2015dynamical,
  zhiqiang2017nonequilibrium, smale2018observation}, and precision
metrology\cite{wineland1992spin, kitagawa1993squeezed, ma2011quantum,
  zhong2010simplified}.  With the development of advanced atomic
trapping, cooling, and control techniques, collective spin systems
have been experimentally realized in atomic
ensembles\cite{takano2009spin, chen2010heralded}, Bose-Einstein
condensates\cite{riedel2010atomchipbased, gross2010nonlinear},
ultracold Fermi gasses\cite{martin2013quantum}, trapped
ions\cite{bohnet2016quantum}, optical cavities\cite{hosten2016quantum,
  norcia2018cavitymediated}, and molecular
sensors\cite{jones2009magnetic}.

One of the primary motivations for studying collective spin systems is
their application to quantum-enhanced metrology.  Quantum projection
noise limits the error $\Delta\phi$ in the measurement of a phase
angle $\phi$ with $N$ independent spins to
$\Delta\phi\sim1/\sqrt{N}$\cite{wineland1992spin, itano1993quantum,
  ma2011quantum}.  Collective spin systems enable a means to break
through this limit via the preparation of many-body entangled states
such as NOON\cite{jones2009magnetic, chen2010heralded,
  zhong2010simplified}, spin-cat\cite{huang2015quantum,
  agarwal1997atomic, lau2014proposal}, and spin-squeezed
states\cite{takano2009spin, riedel2010atomchipbased,
  gross2010nonlinear, hosten2016quantum, norcia2018cavitymediated,
  wineland1992spin, kitagawa1993squeezed, ma2011quantum} that allow
for measurement errors $\Delta\phi\sim1/N^\epsilon$ with
$\epsilon>1/2$.  Generating such states requires nonlinear dynamics
that are typically realized using long-range collisional, photon-, or
phonon-mediated interactions.
% typically resulting in a one-axis twising (OAT) model described by
% the Hamiltonian\cite{kitagawa1993squeezed}
% \begin{align}
%   H_{\t{OAT}} = \chi S_\z^2,
%   \label{eq:OAT}
% \end{align}
% where $S_\z\equiv\sum_{j=1}^N\sigma_\z^{(j)}/2$ is a collective
% spin-$z$ operator defined in terms of Pauli-$z$ operators
% $\sigma_\z^{(j)}$ addressing spin $j$.  The OAT model prepares a
% spin-squeezed state that allows measurement of a phase angle $\phi$
% with error $\Delta\phi\sim N^{-2/3}$ at times
% $t\sim\chi^{-1}N^{-2/3}$\cite{ma2011quantum}.
Although a truly collective spin model requires uniform, all-to-all
interactions, as long as measurements do not distinguish between
constituent particles even non-uniform systems can be effectively
described by a uniform model with renormalized
parameters\cite{hu2015entangled}.
% Various works have proposed improvements over OAT through the
% addition of auxiliary control fields\cite{law2001coherent,
% liu2011spin, huang2015twoaxis, muessel2015twistandturn} or a
% modified vacuum\cite{hu2017vacuum}, thereby realizing different
% collective spin models.

% While the OAT model admits exact
% solutions\cite{foss-feig2013nonequilibrium}, the same is not true of
% more general, non-Ising Hamiltonians.  Nonetheless,
In the absence of decoherence, permutation symmetry and total spin
conservation divide the total Hilbert space of a collective spin
system into superseletion sectors that grow only linearly with system
size $N$, thereby admitting efficient classical simulation of its
dynamics.  Decoherence will generally violate total spin conservation
and require the use of density operators, increasing the dimension of
accessible state space to $\sim N^3$\cite{hartmann2016generalized,
  xu2013simulating} and allowing for exact simulations of
$N\lesssim100$ particles.  If decoherence is sufficiently weak,
dynamics are numerically solvable for $N\lesssim10^4$ particles via
``quantum jump'' Monte Carlo methods\cite{plenio1998quantumjump,
  zhang2018montecarlo} (also known as ``quantum trajectory'' or
``Monte Carlo wavefunction'' methods) that can reproduce all
expectation values of interest.  When decoherence is strong, however,
or when the number of jump operators grows extensively with system
size (e.g.~for single-spin decay), these Monte Carlo methods can take
a prohibitively long time to converge, as simulations become dominated
by incoherent jumps that generate large numbers of distinct
trajectories over which to average.

In this work, we develop a new method to simulate short-time dynamics
of collective spin systems with permutationally-symmetric decoherence.
This method is based on a short-time expansion of exact solutions to
the equations of motion for Heisenberg operators.  Evaluating this
expansion requires use of the structure factors of a collective spin
operator algebra; the calculation of these structure factors is one of
the main technical results of this work, which we hope will enable
future studies to develop additional analytical insights into the
structure and dynamics of collective spin systems.

@@

The time derivative operator $d/dt$ will generally couple spin
operators $\S_{\v m}$ to spin operators $\S_{\v n}$ with higher
``weight'', i.e.~with $\abs{\v n}>\abs{\v m}$, where
$\abs{\v\ell}\equiv\sum_\lambda\ell_\lambda$.  The growth of operator
weight signifies the growth of many-body correlations.  In practice,
keeping track of this growth will eventually require more
computational resources than are available, meaning we must somehow
truncate our equations of motion.  The simplest truncation strategy
would be to take
\begin{align}
  \f{d}{dt} \S_{\v m}
  \to \sum_{w\p{\v n}<W} T_{\v m\v n} \S_{\v n}
  \label{eq:weight_truncation}
\end{align}
for some weight measure $w$, e.g.~$w\p{\v n}=\abs{\v n}$, and cutoff
$W$.  The truncation in Eq.~\eqref{eq:weight_truncation} closes the
system of differential equations defined by Eq.~\eqref{eq:EOM}, and
allow us to solve it using standard numerical methods.  Initial
conditions for this system of differential equations, i.e.~expectation
values of collective spin operators with respect to states that are
simple to prepare experimentally, are provided in Appendix
\ref{sec:initial_conditions}.

The truncation strategy in Eq.~\eqref{eq:weight_truncation} has a few
limitations:
\begin{enumerate*}[label=(\roman*)]
\item simulating a system of differential equations for a large number
  of operators can be quite time-consuming,
\item the weight measure $w$ may need to be chosen carefully, as the
  optimal measure is generally system-dependent,
\item simulation results can only be trusted up to the time at which
  the initial value of operators $\S_{\v m}$ with weight
  $w\p{\v m}\ge W$ have a non-negligible contribution to expectation
  values of interest.\label{pt:limitation}
\end{enumerate*}
The last limitation in particular unavoidably applies in some form to
any method tracking only a subset of all relevant operators.  We
therefore devise an alternate truncation strategy built around
limitation \ref{pt:limitation}.

@@

\section{Spin squeezing}

@@

Initial conditions for all collective spin operators with respect to
(spin-polarized) Gaussian states in the Dicke manifold spanned by
$\set{\ket{m}:S_\z\ket{m}=m\ket{m}}$ can be determined from the
expansions
\begin{align}
  S_\z = \sum_{k=-S}^S k \op{k},
  &&
  S_\mu = \sum_{k=-S+\delta_{\mu,-1}}^{S-\delta_{\mu,1}}
  g_\mu\p{S,k} \op{k+\mu}{k},
\end{align}
with $\mu\in\set{+1,-1}$ and
\begin{align}
  g_\mu\p{S,k} \equiv \sqrt{\p{S-\mu k}\p{S+\mu k+1}}.
\end{align}
Given the initial states
\begin{align}
  \ket{\pm\Z} \equiv \ket{\pm S},
  &&
  \ket{\X} \equiv 2^{-N/2} \sum_{k=-S}^S
  { N \choose S+k }^{1/2} \ket{k},
\end{align}
where $S_\x\ket{\X}=S\ket{\X}$, in terms of $\nu\in\set{+1,-1}$ and
$\bmu\equiv-\mu$ we then have
\begin{align}
  \bk{\nu\Z|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\nu\Z}
  = \delta_{\ell n} \times
  \begin{cases}
     \p{S-n}^m \f{N! n!}{\p{N-n}!} & \mu = \nu, \\
     \delta_{n,0} \p{-S}^m & \mu \ne \nu,
  \end{cases}
\end{align}
and
\begin{align}
  \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}
  &=  \f1{2^N} \sum_{k=-S+\delta_{\mu,-1}\max\set{\ell,n}}
  ^{S-\delta_{\mu,1}\max\set{\ell,n}} \p{\mu k}^m
  \sp{{ N \choose S+k+\mu\ell } { N \choose S+k+\mu n }}^{1/2}
  \notag \\
  &\qquad\qquad\qquad\qquad
  \times \prod_{p=0}^{\ell-1} g_\mu\p{S,k+\mu p}
  \prod_{q=0}^{n-1} g_\bmu\p{S,k+\mu\sp{n-q}} \\
  &= \f{N!}{2^N} \sum_{\mu k=-\mu S-\delta_{\mu,-1}\max\set{\ell,n}}
  ^{\mu S-\delta_{\mu,1}\max\set{\ell,n}}
  \f{\p{\mu k}^m \p{S-\mu k}!}
  {\p{S+\mu k}!\p{S-\ell-\mu k}!\p{S-n-\mu k}!} \\
  &= \f{N!}{2^N} \sum_{k=-S}^{S-\max\set{\ell,n}}
  \f{k^m \p{S-k}!}{\p{S+k}!\p{S-\ell-k}!\p{S-n-k}!} \\
  &= \p{-1}^m\f{N!}{2^N} \sum_{k=0}^{N-\max\set{\ell,n}}
  \f{\p{S-k}^m \p{N-k}!}{k!\p{N-\ell-k}!\p{N-n-k}!}.
\end{align}
Furthermore, using the fact that
\begin{align}
  e^{i\phi S_\z} S_\mu e^{-i\phi S_\z} = e^{i\mu\phi} S_\mu,
\end{align}
we can define the state
\begin{align}
  \ket{\tilde\phi} \equiv e^{-i\phi S_\z}\ket{\X},
  &&
  \p{\cos\phi S_\x + \sin\phi S_\y} \ket{\tilde\phi}
  = S \ket{\tilde\phi},
\end{align}
and compute
\begin{align}
  \bk{\tilde\phi|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\tilde\phi}
  = e^{i\mu\p{\ell-n}\phi} \bk{\X|S_\mu^\ell \p{\mu S_\z}^m S_\bmu^n|\X}.
\end{align}

@@

$P_{\v j}$ in \eqref{eq:S_P} we define
